---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Unzip and load the data in a data frame

```{r}
unzip("activity.zip")
data <- read.csv("activity.csv",stringsAsFactors = FALSE)
data$date <- as.Date.character(data$date)
```


## What is mean total number of steps taken per day?

splitting the data frame on date, vectorized sum and plotting

```{r}
data_day <- split(data, data$date)
steps_day_total <- sapply(data_day, function(x) {sum(x$steps)})
hist(steps_day_total, xlab = "Steps", main = "Histogram of total # of steps per day")
```

Mean and median number of steps taken each day

```{r}
steps_day_mean <- sapply(data_day, function(x) {mean(x$steps)})
steps_day_median <- sapply(data_day, function(x) {median(x$steps)})
```

Note: the median consists of zeros and NAs, meaning that more than half of the day no steps were taken

```{r}
steps_day_median
```

Plotting the time series of average take per day

```{r}
days_scale <- sapply(data_day, function(x) {max(x$date)})
time_serie <- cbind(days_scale, steps_day_mean)
plot(time_serie[,1], time_serie[,2], type = "l", xlab = "days", ylab = "average steps", main = "Average # steps per interval by day")
```

## What is the average daily activity pattern?

Split the data on interval

```{r}
data_interval <- split(data, data$interval)
steps_interval_mean <- sapply(data_interval, function(x) {mean(x$steps, na.rm = TRUE)})
```

Plot the profile during the day

```{r}
plot(as.numeric(names(steps_interval_mean)),steps_interval_mean, type = "l", main = "# of steps per interval", xlab = "", ylab = "", ylim = c(0,200))
```

Find the max value name and its index and the corresponding number of steps value

```{r}
index <- which.max(steps_interval_mean)
index
steps_interval_mean[index]
```



## Imputing missing values

The strategy that will be used it to replace NAs with the interval average
The new data set will be data2

```{r}
data2 <- data
for (i in 1:length(data$steps)) {
    if (is.na(data$steps[i])) {
        ref <- match(data$interval[i], names(steps_interval_mean))
        data2$steps[i] <- steps_interval_mean[ref]
    }
}
```

Revised histogram: the frequency of the center of the distribution is higher since we have replaced not available data with average data

```{r}
data_day2 <- split(data2, data2$date)
steps_day_total2 <- sapply(data_day2, function(x) {sum(x$steps)})
hist(steps_day_total2,xlab = "Steps", main = "Histogram of total # of steps per day after replacing missing values")
```

## Are there differences in activity patterns between weekdays and weekends?

Creating weekday boolean and splitting the data

```{r}
data$weekdays <- ifelse(weekdays(data$date, abbreviate = TRUE)=="Sat" | weekdays(data$date, abbreviate = TRUE)=="Sun",0,1)

data_interval_wday <- split(data[data$weekdays==1,], data[data$weekdays==1,3])
steps_interval_mean_wday <- sapply(data_interval_wday, function(x) {mean(x$steps, na.rm = TRUE)})

data_interval_wkend <- split(data[data$weekdays==0,], data[data$weekdays==0,3])
steps_interval_mean_wkend <- sapply(data_interval_wkend, function(x) {mean(x$steps, na.rm = TRUE)})
```

Plotting the results: activity starts later on week ends, but is higher during the usual 'working hours' on week ends, due to not being stuck at one's desk likely

```{r, fig.height=10}
par(mfrow=c(2,1))
plot(as.numeric(names(steps_interval_mean_wkend)),steps_interval_mean_wkend, type = "l", main = "# of steps per interval on weekend", xlab = "", ylab = "", ylim = c(0,200))
plot(as.numeric(names(steps_interval_mean_wday)),steps_interval_mean_wday, type = "l", main = "# of steps per interval on weekday",xlab = "", ylab = "", ylim = c(0,200))

```
